Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-02-23 newpaper/paper.tex
lots of stream fusion fiddling

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}

\newcommand{\SOP}{\msf{SOP}}
\newcommand{\El}{\msf{El}}
\newcommand{\USOP}{\msf{U}_{\msf{SOP}}}
\newcommand{\Uprod}{\msf{U_P}}
\newcommand{\Elprod}{\msf{El_{P}}}
\newcommand{\IsSOP}{\msf{IsSOP}}
\newcommand{\forEach}{\msf{forEach}}
\newcommand{\single}{\msf{single}}
\newcommand{\msplit}{\msf{split}}
\newcommand{\mapGen}{\msf{mapGen}}

@@

We demonstrate this for lists first. An object-level $\mbf{case}$ on lists introduces
two points where code generation can continue. We define a metatype which gives
us a ``view'' on these points:
\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
We can generate code for a case split, returning a view on it:
\begin{alignat*}{3}
  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{a}\,\qt{\vas}))}
\end{alignat*}
Now, in any $\MonadGen$, we may write
\begin{alignat*}{3}
  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
\end{alignat*}
%% \begin{alignat*}{3}
%%   &\mdo && \mit{sp} \fro \liftGen \dlr \mit{split}\,\vas\\
%%   &     && \case \mit{sp}\,\of \\
%%   &     && \ind \Nil' \to ... \\
%%   &     && \ind \Cons'\,a\,\vas \to ...
%% \end{alignat*}
This can be generalized to splitting on any object value. In the supplemental
Haskell and Agda implementations, we overload $\mit{split}$ with a type
class. However, in a native implementation of CFTT it might make sense to extend
$\mdo$-notation, so that we elaborate $\mbf{case}$ on object values to an
application of the appropriate $\mit{split}$ function. We adopt this in the rest
of the paper, so we will be able to write $\case \vx\,\of ...$ in a $\mdo$-block,
whenever we work in a $\MonadGen$ and $\vx$ has a value type.

@@

%% We implemented all features in this section in typed Template Haskell
%% \cite{TODO}, using only the strongly typed interface. The organization of the
%% code is essentially the same as here, but some changes are necessary:
%% \begin{itemize}
%%   \item TH cannot properly handle scoping of type variables across quotations,
%%         generating ill-typed code in some cases. It is fairly easy to work around
%%         this using term-level proxy values.
%%   \item We do not bother tracking value and computation types. This could be done,
%%         using dummy type classes, but we would not get any extra guarantees from it
%%         in downstream compilation.
%%   \item When we take the product of generic sums in CFTT, we simply take the
%%         object-level product of component types, pairwise. We do this because we
%%         assume that the downstream compiler can represent products without
%%         overheads, in an unboxed manner, and also erase unit types. In Haskell we do
%%         not rely on this, and instead use a slightly more complicated
%%         sum-of-products-of-values representation, and also fuse away product types
%%         during staging.
%%   \item We make heavy use of \emph{singleton types} \cite{TODO} in generic sums-of-products, to emulate
%%         dependent types. This adds significant noise compared to a native dependently typed
%%         version.
%%   \item We have two separate type classes for binding-time improvement, one for
%%         base types and one for monads specifically. This is because the quantified
%%         constraints in \ref{TODO} yield poor type inference for $\mup$ and $\mdown$,
%%         by preventing \emph{functional dependencies} \cite{TODO} between the class
%%         parameters. We expect that in a native CFTT implementation, the extra staging
%%         information in type universes would help inference in this case.
%% \end{itemize}

@@

%% Fusion techniques can be broadly grouped into \emph{push} fusion, which is based
%% on Church-encodings of inductive lists, and $\emph{pull}$ fusion, which is based
%% on Church-encodings of coinductive lists, which manifests as \emph{state
%% machines} \cite{TODO}. The two styles have different trade-offs, and in
%% practical programming it is a good idea to support both, but in this section we
%% focus on pull fusion.

%% The reason is partly lack of space, but also that pull fusion has been
%% historically the more difficult to reliably compile, and we can demonstrate
%% significant improvements in CFTT, in robustness and expressive power. Our fusion
%% always succeeds, and we also support a $\mit{concatMap}$ operation, which
%% has been elusive in guaranteed fusion. We believe that our implementation also
%% provides valuable conceptual clarity.

@@

A pull stream is a meta-level specification of a state machine:
\begin{alignat*}{3}
  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
\end{alignat*}
In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
constraint. The next field with type $S$ is the initial state, while transitions
are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$ field. Possible transitions
are stopping ($\Stop$), transitioning to a new state while outputting a value
($\Yield$) and making a silent transition to a new state ($\Skip$).

Let us see some operations on streams now. First, $\Pull$ is a ``zippy''
applicative functor, where $\mit{pure}$ infinitely repeats a value and $({\ap})$
applies a stream of functions to a stream of values, pointwise.
\begin{alignat*}{3}
  &\mit{pure} : \vA \to \Pull\,\vA\\
  &\mit{pure} = \Pull\,()\,()\,(\lam \_.\,\return \dlr \Stop)
\end{alignat*}
\begin{alignat*}{3}
  &(\!\ap\!) : \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB \\
  &(\!\ap\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =\\
  & \ind \Pull\,(\vS,\,\vS')\,(\seed,\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
  & \ind \ind \Stop \hspace{1.20em}        \to \return \Stop \\
  & \ind \ind \Skip\,\vs \hspace{0.75em}     \to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \Yield\,\vf\,\vs \to \step'\,\vs' \bind \lam \case \\
  & \ind \ind \ind \Stop \hspace{1.7em} \to \return \Stop \\
  & \ind \ind \ind \Skip\,\vs'\hspace{0.9em}\to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \ind \Yield\,\va\,\vs' \to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs')
\end{alignat*}
In $\mit{pure}$, the state is the unit type, while in $(\!\ap\!)$ we take the
product of the stream states and do synchronous transitions. In both
definitions, the $\IsSOP\,\vS$ constraint is implicitly dispatched by instance
resolution; this works in the Agda and Haskell versions too. We can derive the
usual $\mit{zip}$ and $\mit{zipWith}$ operations from $(\!\ap\!)$. $\Pull$
is also a monoid with stream appending as the binary operation.
\begin{alignat*}{3}
  & \mit{empty} : \Pull\,\vA\\
  & \mit{empty} = \Pull\,()\,()\,(\lam \_.\,\return \Stop)
\end{alignat*}
\begin{alignat*}{5}
  & \hspace{-4em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
  & \hspace{-4em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left\,\seed)\,\lam \case} \\
  & \hspace{-4em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to \return \dlr \Skip\,(\Right\,\seed')\\
  & \hspace{-4em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
  & \hspace{-4em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
  & \hspace{-4em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
  & \hspace{-4em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
\end{alignat*}
These definitions are standard for streams; note though that compared to the
unstaged definitions in previous literature, the only additional noise is just
the $\Gen$ monad in the transitions. Likewise, we can give standard definitions
for usual stream functions such as $\msf{filter}$, $\msf{take}$ or
$\msf{drop}$.

@@

We redefine the previous $\mit{Fun}_{\SOP\Lift}$ to return a computation type
instead:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \CTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\mit{foldr}\,(\to)\,\vR\,\vA,\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR)\\
  &\mindex    &&: \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR) \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
For join points, this is just as efficient as the previous
$\mit{Fun}_{\SOP\Lift}$, since a definition of a product of functions gets
compiled to a sequence of function definitions. Now, we can generate object code
from streams. There are several choices for this, but in CFTT the $\msf{foldr}$
function is as good as we can get.
\begin{alignat*}{3}
  &\hspace{-13em}\mathrlap{\foldr : \{\vA : \MTy\}\{\vB : \Ty\} \to (\vA \to \Lift \vB \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB}\\
  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,(\Pull\,\vS\,\seed\,\step)\,\vf\,\vb = \ql}\\
  &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
  &\hspace{-13em}\ind \ind \Stop            &&\to \vb \\
  &\hspace{-13em}\ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
  &\hspace{-13em}\ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
  &\hspace{-13em}\mathrlap{\ind \spl(\mindex\,\qt{\vfs}\,(\rep\,\seed))\qr}
\end{alignat*}
This $\foldr$ is quite flexible, becuse we can eliminate into any object type,
including function types. For instance, we can define $\foldl$ from $\foldr$:
\begin{alignat*}{3}
  &\foldl : \{\vA : \MTy\}\{\vB : \VTy\} \to (\Lift \vB \to \vA \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB\\
  &\foldl\,\vf\,\vb\,\vas = \spl(\foldr\,(\lam \va\,\vg.\,\qt{\lam \vb.\,\spl \vg\,\spl(\vf\,\qt{\vb}\,\spl \va)})\,\qt{\lam \vb.\,\vb}\,\vas)\,\spl \vb
\end{alignat*}
Note that since we abstract $\vB$ in a runtime function, it must be a value
type. Here, each $\Stop$ and $\Yield$ in the transition function gets
interpreted as a $\lambda$-expression in the output. However, those $\lambda$-s
will be lifted out in the scope, yielding a proper mutually tail-recursive
definition with an accummulator for $\vB$. Contrast this to the GHC base
library, where $\foldl$ for lists is also defined from $\foldr$, to enable
push-based fusion, but where a substantial \emph{arity analysis} is required in
the compiler to eliminate the intermediate closures \cite{TODO}.

@@

\subsection{Let-Insertion and Case Splitting}

Since $\Pull$ is not a monad, it cannot be $\MonadGen$ in our framework,
but we still support some forms of let-insertion. First, we note that $\Pull$ can be
mapped over by functions returning in $\Gen$:
\begin{alignat*}{3}
  &\hspace{-3em}\mathrlap{\mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB }\\
  &\hspace{-3em}\mathrlap{\mit{mapGen}\,\vf\,(\Pull\,\vS\,\seed\,\step) = \Pull\,\vS\,\seed \dlr \lam \case}\\
  &\hspace{-3em} \ind \Stop            &&\to \return\,\Stop \\
  &\hspace{-3em} \ind \Skip\,\vs       &&\to \return \dlr \Skip\,\vs\\
  &\hspace{-3em} \ind \Yield\,\va\,\vs &&\to \mdo \{\vb \fro \vf\,\va; \return \dlr \Yield\,\vb\,\vs\}
\end{alignat*}
Now, we may use $\mit{mapGen}\,\gen : \Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA)$
to let-bind all yielded expressions in a stream. This can be useful in stream
operations that would otherwise duplicate yielded expressions. For example, the
function with type $\Pull\,(\Lift \vA) \to \Pull\,(\Lift \vA,\,\Lift \vA)$
should be defined with let-insertion.

Let us now consider a more ``monadic'' style of stream programming. For this we
need singleton streams:
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
We also write $\mit{forEach}$ for $\concatMap$ with flipped arguments. Now, a
more monadic flavor of let-insertion is given as
\begin{alignat*}{3}
  &\gen_{\Pull} : \Lift \vA \to \Pull\,(\Lift \vA) \\
  &\gen_{\Pull}\,\va = \mit{mapGen}\,\gen\,(\mit{single}\,\va).
\end{alignat*}
We can also implement case splitting on object values. Recall from Section
\ref{TODO} that for each value type we have a splitting function that returns a
meta-level view on the constructors. Importantly, all such views are $\SOP$-s, and
therefore can be bound by $\forEach$. A small example:
\begin{alignat*}{3}
  & \hspace{-5em}\mathrlap{\forEach\,(\mit{countFrom}\,\qt{0}) \dlr \lam \vn. }\\
  & \hspace{-5em}\mathrlap{\forEach\,(\gen_{\Pull}\,\qt{\spl \vn * 2}) \dlr \lam \vm. }\\
  & \hspace{-5em}\mathrlap{\forEach\,(\single\,(\msplit\,\qt{\spl \vm < 20})) \dlr \lam \case }\\
  & \hspace{-5em}\ind \True_\M  &&\to \single\,\qt{\spl \vm + 10}   \\
  & \hspace{-5em}\ind \False_\M &&\to \single\,\qt{\spl \vm + 20}
\end{alignat*}
Here $(\gen_{\Pull}\,\qt{\spl \vn * 2})$ behaves like a let-insertion, and
$(\single\,(\msplit\,\qt{\spl \vm < 20}))$ implements the case splitting.

>>>>

\newcommand{\vas}{\mathsf{as}}
\newcommand{\vbs}{\mathsf{bs}}
\newcommand{\vcs}{\mathsf{cs}}
\newcommand{\vxs}{\mathsf{xs}}
\newcommand{\vys}{\mathsf{ys}}
\newcommand{\vsp}{\mathsf{sp}}
\newcommand{\vma}{\mathsf{ma}}
\newcommand{\vm}{\mathsf{m}}
\newcommand{\vn}{\mathsf{n}}
\newcommand{\vk}{\mathsf{k}}
\newcommand{\vA}{\mathsf{A}}
\newcommand{\vB}{\mathsf{B}}
\newcommand{\vC}{\mathsf{C}}
\newcommand{\vS}{\mathsf{S}}
\newcommand{\vF}{\mathsf{F}}
\newcommand{\vR}{\mathsf{R}}
\newcommand{\vM}{\mathsf{M}}
\newcommand{\vmb}{\mathsf{mb}}
\newcommand{\mAs}{\mathsf{As}}
\newcommand{\va}{\mathsf{a}}
\newcommand{\vb}{\mathsf{b}}
\newcommand{\vc}{\mathsf{c}}
\newcommand{\vd}{\mathsf{d}}
\newcommand{\vx}{\mathsf{x}}
\newcommand{\vy}{\mathsf{y}}
\newcommand{\vz}{\mathsf{z}}
\newcommand{\vf}{\mathsf{f}}
\newcommand{\vfs}{\mathsf{fs}}
\newcommand{\vg}{\mathsf{g}}
\newcommand{\vh}{\mathsf{h}}
\newcommand{\vt}{\mathsf{t}}
\newcommand{\vs}{\mathsf{s}}
\newcommand{\vr}{\mathsf{r}}
\newcommand{\vu}{\mathsf{u}}

\newcommand{\SOP}{\msf{SOP}}
\newcommand{\El}{\msf{El}}
\newcommand{\USOP}{\msf{U}_{\msf{SOP}}}
\newcommand{\Uprod}{\msf{U_P}}
\newcommand{\Elprod}{\msf{El_{P}}}
\newcommand{\IsSOP}{\msf{IsSOP}}
\newcommand{\forEach}{\msf{forEach}}
\newcommand{\single}{\msf{single}}
\newcommand{\msplit}{\msf{split}}
\newcommand{\mapGen}{\msf{mapGen}}
\newcommand{\genPull}{\msf{gen_{Pull}}}
\newcommand{\casePull}{\msf{case_{Pull}}}

@@

We demonstrate this for lists first. An object-level $\mbf{case}$ on lists introduces
two points where code generation can continue. We define a metatype which gives
us a ``view'' on these points:
\[ \data \msf{SplitList}\,\vA = \Nil'\,|\,\Cons'\,(\Up \vA)\,(\Up (\List\,\vA)) \]
We can generate code for a case split, returning a view on it:
\begin{alignat*}{3}
  &\mit{split} : \Up (\List\,\vA) \to \Gen\,(\msf{SplitList}\,\vA)\\
  &\mit{split}\,\vas = \Gen \dlr \lam \vk.\,\qt{\case \spl \vas\,\of \Nil \to \spl(\vk\,\Nil'); \Cons\,\va\,\vas \to \spl(\vk\,(\Cons'\,\qt{a}\,\qt{\vas}))}
\end{alignat*}
Now, in any $\MonadGen$, we may write
\begin{alignat*}{3}
  &\mdo \{\mit{sp} \fro \liftGen\;(\!\mit{split}\,\vas);(\case \mit{sp}\,\of \Nil' \to ...;\Cons'\,\va\,\vas \to ...)\}
\end{alignat*}
This can be generalized to splitting on any object value. In the Agda and
Haskell implementations, we overload $\mit{split}$ with a class similar to the
following:
\begin{alignat*}{3}
  & \hspace{-2em}\mathrlap{\tyclass\,\mit{Split}\,(\vA : \VTy)\,\where}\\
  & \hspace{-2em}\ind \mit{SplitTo} &&: \MTy \\
  & \hspace{-2em}\ind \mit{split}   &&: \Lift \vA \to \Gen\,\mit{SplitTo}
\end{alignat*}
In a native implementation of CFTT it may make sense to extend $\mdo$-notation,
so that we elaborate $\mbf{case}$ on object values to an application of the
appropriate $\mit{split}$ function. We adopt this in the rest of the paper, so
we will be able to write $\case \vx\,\of ...$ in a $\mdo$-block, whenever we
work in a $\MonadGen$ and $\vx$ has a value type.

@@

@@

@@

A pull stream is a meta-level specification of a state machine:
\begin{alignat*}{3}
  & \data \Step\,\vS\,\vA = \Stop\,|\,\Skip\,\vS\,|\,\Yield\,\vA\,\vS\\
  & \data \Pull\,(\vA : \MTy) : \MTy\,\where\\
  & \ind \Pull : (\vS : \MTy) \to \IsSOP\,\vS \RA \Gen\,\vS \to (\vS \to \Gen\,(\Step\,\vS\,\vA)) \to \Pull\,\vA
\end{alignat*}
In the $\Pull$ constructor, $\vS$ is the type of the internal state which is
required to be a sum-of-products of value types by the $\IsSumVS\,\vS$
constraint. The next field with type $\Gen\,S$ is the initial state, while transitions
are represented by the $\vS \to \Gen\,(\Step\,\vS\,\vA)$ field. Possible transitions
are stopping ($\Stop$), transitioning to a new state while outputting a value
($\Yield$) and making a silent transition to a new state ($\Skip$).

Let us see some operations on streams now. First, $\Pull$ is a ``zippy''
applicative functor, where $\mit{pure}$ infinitely repeats a value and $({\ap})$
applies a stream of functions to a stream of values, pointwise.
\begin{alignat*}{3}
  &\mit{pure} : \vA \to \Pull\,\vA\\
  &\mit{pure}\,\va = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \dlr \Yield\,\va\,())
\end{alignat*}
\begin{alignat*}{3}
  &(\!\ap\!) : \Pull\,(\vA \to \vB) \to \Pull\,\vA \to \Pull\,\vB \\
  &(\!\ap\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') =\\
  & \ind \Pull\,(\vS,\,\vS')\,(({,}) \fmap \seed\,\ap\,\seed') \dlr \lam (\vs,\,\vs').\,\step\,\vs \bind \lam \case \\
  & \ind \ind \Stop \hspace{1.20em}        \to \return \Stop \\
  & \ind \ind \Skip\,\vs \hspace{0.75em}     \to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \Yield\,\vf\,\vs \to \step'\,\vs' \bind \lam \case \\
  & \ind \ind \ind \Stop \hspace{1.7em} \to \return \Stop \\
  & \ind \ind \ind \Skip\,\vs'\hspace{0.9em}\to \return \dlr \Skip\,(\vs,\,\vs') \\
  & \ind \ind \ind \Yield\,\va\,\vs' \to \return \dlr \Yield\,(\vf\,\va)\,(\vs,\,\vs')
\end{alignat*}
In $\mit{pure}$, the state is the unit type, while in $(\!\ap\!)$ we take the
product of the stream states and do synchronous transitions. In both
definitions, the $\IsSOP\,\vS$ constraint is implicitly dispatched by instance
resolution; this works in the Agda and Haskell versions too. We can derive the
usual $\mit{zip}$ and $\mit{zipWith}$ operations from $(\!\ap\!)$. $\Pull$
is also a monoid with stream appending as the binary operation.
\begin{alignat*}{3}
  & \mit{empty} : \Pull\,\vA\\
  & \mit{empty} = \Pull\,()\,(\return\,())\,(\lam \_.\,\return \Stop)
\end{alignat*}
\begin{alignat*}{5}
  & \hspace{-5em}\mathrlap{(\!<>\!) : \Pull\,\vA \to \Pull\,\vA \to \Pull\,\vA} \\
  & \hspace{-5em}\mathrlap{(\!<>\!)\,(\Pull\,\vS\,\seed\,\step)\,(\Pull\,\vS'\,\seed'\,\step') = \Pull\,(\Either\,\vS\,\vS')\,(\Left \fmap \seed)\,\lam \case} \\
  & \hspace{-5em}\ind \Left\,\vs &&\to \step\,s \bind \lam \case &&\Stop             &&\to (\Skip \circ \Right) \fmap \seed'\\
  & \hspace{-5em}                    &&                          &&\Skip\,\vs        &&\to \return \dlr \Skip\,(\Left\,\vs) \\
  & \hspace{-5em}                    &&                        &&\Yield\,\va\,\vs    &&\to \return \dlr \Yield\,\va\,(\Left\,\vs)\\
  & \hspace{-5em}\ind \Right\,\vs' &&\to \step\,s' \bind \lam \case &&\Stop          &&\to \return \Stop\\
  & \hspace{-5em}                    &&                        &&\Skip\,\vs'         &&\to \return \dlr \Skip\,(\Right\,\vs') \\
  & \hspace{-5em}                    &&                        &&\Yield\,\va\,\vs'   &&\to \return \dlr \Yield\,\va\,(\Right\,\vs')
\end{alignat*}
These definitions are standard for streams; note though that compared to the
unstaged definitions in previous literature, the only additional noise is just
the $\Gen$ monad in the initial states and the transitions. Likewise, we can
give standard definitions for usual stream functions such as $\msf{filter}$,
$\msf{take}$ or $\msf{drop}$.

@@

We redefine the previous $\mit{Fun}_{\SOP\Lift}$ to return a computation type
instead:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \CTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\mit{foldr}\,(\to)\,\vR\,\vA,\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR)\\
  &\mindex    &&: \Lift(\mit{Fun}_{\SOP\Lift}\,\vA\,\vR) \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
Even for join points, this is just as efficient as the previous
$\mit{Fun}_{\SOP\Lift}$ version, since a definition of a product of functions
gets compiled to a sequence of function definitions. Now, we can generate object
code from streams. There are several choices for this, but in CFTT the
$\msf{foldr}$ function is as good as we can get.
\begin{alignat*}{3}
  &\hspace{-13em}\mathrlap{\foldr : \{\vA : \MTy\}\{\vB : \Ty\} \to (\vA \to \Lift \vB \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB}\\
  &\hspace{-13em}\mathrlap{\foldr\,\{\vA\}\,\{\vB\}\,(\Pull\,\vS\,\seed\,\step)\,\vf\,\vb = \ql}\\
  &\hspace{-13em}\mathrlap{\ind \letrec \vfs : \mit{Fun}_{\SOP\Lift}\,(\Rep\,\{\vS\})\,\vB := \spl\big(\mtabulate \dlr \lam \vs.\,\unGen\,(\step\,(\rep^{-1}\,\vs)) \dlr \lam \case} \\
  &\hspace{-13em}\ind \ind \Stop            &&\to \vb \\
  &\hspace{-13em}\ind \ind \Skip\,\vs       &&\to \mindex\,\qt{\vfs}\,(\rep\,\vs) \\
  &\hspace{-13em}\ind \ind \Yield\,\va\,\vs &&\to \vf\,\va\,(\mindex\,\qt{\vfs}\,(\rep\,\vs))\big);\\
  &\hspace{-13em}\mathrlap{\ind \spl\big(\runGen \dlr \mdo \{s \fro \seed; \return \dlr \mindex\,\qt{\vfs}\,(\rep\,\vs)\}\big)\qr}
\end{alignat*}
This $\foldr$ is quite flexible, becuse we can eliminate into any object type,
including function types. For instance, we can define $\foldl$ from $\foldr$:
\begin{alignat*}{3}
  &\foldl : \{\vA : \MTy\}\{\vB : \VTy\} \to (\Lift \vB \to \vA \to \Lift \vB) \to \Lift \vB \to \Pull\,\vA \to \Lift\,\vB\\
  &\foldl\,\vf\,\vb\,\vas = \spl(\foldr\,(\lam \va\,\vg.\,\qt{\lam \vb.\,\spl \vg\,\spl(\vf\,\qt{\vb}\,\spl \va)})\,\qt{\lam \vb.\,\vb}\,\vas)\,\spl \vb
\end{alignat*}
Note that since we abstract $\vB$ in a runtime function, it must be a value
type. Here, each $\Stop$ and $\Yield$ in the transition function gets
interpreted as a $\lambda$-expression in the output. However, those $\lambda$-s
will be lifted out in the scope, yielding a proper mutually tail-recursive
definition with an accummulator for $\vB$. Contrast this to the GHC base
library, where $\foldl$ for lists is also defined from $\foldr$, to enable
push-based fusion, but where a substantial \emph{arity analysis} is required in
the compiler to eliminate the intermediate closures \cite{TODO}.

@@

\subsection{Let-Insertion \& Case Splitting in Monadic Style}

While $\Pull$ is not a monad, and hence also not a $\MonadGen$, we can still use
a monadic style of stream programming with good ergonomics. First, we need
singleton streams for ``returning'':
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
Now, this operation should be avoided when possible, since it has two states and
can contribute to a blow-up of state size in combination with state-multiplying
operations such as $\mit{zip}$ or $\concatMap$. Both let-insertion and case
splitting could be defined generically from $\single$, $\concatMap$ and the
following operation (definition omitted):
\[ \mit{mapGen} : (\vA \to \Gen\,\vB) \to \Pull\,\vA \to \Pull\,\vB \]
However, $\single$ would introduce superfluous states that way. Instead, we give
more efficient specialized definitions. Let-insertion is as follows:
\begin{alignat*}{3}
  & \hspace{-8em}\mathrlap{\genPull : \{\vA : \VTy\}\{\vB : \MTy\} \to \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB}\\
  & \hspace{-8em}\mathrlap{\genPull\,\va\,\vf = \Pull\,(\Sigma\,(\Lift\,\vA)\,(\mit{projS} \circ \vf))\,(({\va,}) \fmap \mit{projSeed}\,(\vf\,\va)) \dlr \lam \case}\\
  & \hspace{-8em}\ind \Stop            &&\to \return\,\Stop \\
  & \hspace{-8em}\ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs)\\
  & \hspace{-8em}\ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
\end{alignat*}
For case splitting, we have:
\begin{alignat*}{3}
  &\hspace{-12em}\mathrlap{\casePull : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB }\\
  &\hspace{-12em}\mathrlap{\casePull\,\va\,\vf = \Pull\,(\Sigma\,(\mit{SplitTo}\,\{\vA\})\,(\mit{projS} \circ \vf))}\\
  &\hspace{-12em}\mathrlap{    \hspace{7.1em}         (\mdo \{\va' \fro \mit{split}\,\va; \vs \fro \mit{projSeed}\,(\vf\,\va');\return (\va',\,\vs)\}) \dlr \lam \case}\\
  &\hspace{-4em}\mathrlap{  \ind   (\va',\,\vs) \to \mit{projStep}\,(\vf\,\va')\,\vs \bind \lam \case} \\
  &\hspace{-4em} \ind \ind \ind \Stop            &&\to \return \Stop \\
  &\hspace{-4em} \ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va',\,\vs) \\
  &\hspace{-4em} \ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va',\,\vs)
\end{alignat*}
